---
title: "Modèles mixtes"
subtitle: "TP2: `biodivcontinamerdb`"
author: "Florent Chuffart & Laurent Husson"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---






```{r, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=4.5, eval=TRUE, echo=TRUE, results="verbatim", warning=FALSE)
```

# Data

  -  *lon*            : longitude
  -  *lat*            : lattitude
  -  *z*              : altitude
  -  *rug*            : rugosity
  -  *wsm*            : World Strain Map [1, Fig. 2a], a worldwide quantification of the rate at which the surface of the Earth*s crust is currently shortened, sheared or stretched by tectonic forces
  -  *spr_amphibians* : species richness for amphibians
  -  *spr_birds*      : species richness for birds   
  -  *spr_mammals*    : species richness for mammals
  -  *area*           : geographique area
                 
                 
                


```{r db_files, echo=TRUE, results="verbatim"}
biodivcontinamerdb = readRDS("biodivcontinamerdb.rds")
# set.seed(1)
# biodivcontinamerdb = biodivcontinamerdb[sort(sample(1:nrow(biodivcontinamerdb), 30000)), ]
# saveRDS(biodivcontinamerdb, "biodivcontinamerdb.rds")
# table(biodivcontinamerdb$area)
head(biodivcontinamerdb)
dim(biodivcontinamerdb)
```


```{r map, fig.height=6, fig.width=8}
d = biodivcontinamerdb
# layout(cbind(matrix(1, nrow=3, ncol=3), matrix(2:4)), respect=TRUE)
layout(1, respect=TRUE)
plot(  d$lon, d$lat, pch=".", col=c("grey75", "grey25")[(d$wsm>0)+1], main="North, Central and South American west coast" )
points(d$lon, d$lat, pch=".", col=adjustcolor(as.numeric(d$area), alpha.f=.1))
legend("topright", c("wsm==0", "wsm>0"), col=c("grey75", "grey25"), pch=15)
legend("bottomleft", levels(d$area), col=1:length(levels(d$area)), pch=15)

```


```{r, fig.height=3, fig.width=9}
layout(matrix(1:3, nrow=1), respect=TRUE)
plot(d$lat, d$spr_amphibians, xlab="lat", ylab="spr", main="amphibians", pch=".", col=adjustcolor(as.numeric(d$area), alpha.f=.1))
plot(d$lat, d$spr_birds     , xlab="lat", ylab="spr", main="birds"     , pch=".", col=adjustcolor(as.numeric(d$area), alpha.f=.1))
plot(d$lat, d$spr_mammals   , xlab="lat", ylab="spr", main="mammals"   , pch=".", col=adjustcolor(as.numeric(d$area), alpha.f=.1))
```

```{r}
layout(1, respect=TRUE)
plot(  d$lon, d$lat, pch=".", col=c("grey75", "grey25")[(d$wsm>0)+1], main="North, Central and South American west coast" )
tmpd = d[d$area == "sam_central" & d$spr_birds > 200,]
points(tmpd$lon, tmpd$lat, col=2)
tmpd = d[d$area == "sam_central" & d$spr_birds < 200,]
points(tmpd$lon, tmpd$lat, col=4)

# Warning: take care of factors!
# d$area2 = d$area
# d[d$area == "sam_central" & d$spr_birds > 200,]$area2 = "amazonia"
# table(d$area2 , useNA="ifany")

d$area2 = as.character(d$area)
d[d$area == "sam_central" & d$spr_birds > 200,]$area2 = "amazonia"
d$area2 = factor(d$area2, levels=c(levels(d$area), "amazonia"))


layout(1, respect=TRUE)
plot(  d$lon, d$lat, pch=".", col=c("grey75", "grey25")[(d$wsm>0)+1], main="North, Central and South American west coast" )
points(d$lon, d$lat, pch=".", col=adjustcolor(as.numeric(d$area2), alpha.f=.1))
legend("topright", c("wsm==0", "wsm>0"), col=c("grey75", "grey25"), pch=15)
legend("bottomleft", levels(d$area2), col=1:length(levels(d$area2)), pch=15)
```


```{r spr, fig.height=3, fig.width=9}
layout(matrix(1:3, nrow=1), respect=TRUE)
plot(d$lat, d$spr_amphibians, xlab="lat", ylab="spr", main="amphibians", pch=".", col=adjustcolor(as.numeric(d$area2), alpha.f=.1))
plot(d$lat, d$spr_birds     , xlab="lat", ylab="spr", main="birds"     , pch=".", col=adjustcolor(as.numeric(d$area2), alpha.f=.1))
plot(d$lat, d$spr_mammals   , xlab="lat", ylab="spr", main="mammals"   , pch=".", col=adjustcolor(as.numeric(d$area2), alpha.f=.1))
```





# Questions

Does the World Strain Map impact the species richness given the lattitude for birds?








# Models




```{r data, eval=TRUE, echo=TRUE, results="verbatim"}
d$grp = d$area2
# class
class(d) = c("nfnGroupedData", "nfGroupedData", "groupedData", "data.frame")
attributes(d)$formula = formula(spr_birds ~ lat | grp)
```





```{r echo=TRUE, results="verbatim"}
d$abslat = abs(d$lat)
d$wsm01 = (d$wsm != 0)+0
m01 = lm(spr_birds~abslat+wsm01, d)
layout(1, respect=TRUE)
plot(spr_birds~lat, d, pch=".", col=d$grp)
points(d$lat, predict(m01), cex=.3, col=d$wsm01+1)
legend("topleft", c("wsm==0", "wsm>0"), col=1:2, lty=1)
summary(m01)
```

```{r echo=FALSE, results="verbatim"}
m2 = lm(spr_birds~(abslat+wsm01)*grp, d)
summary(m2)
layout(matrix(1:2, 1, byrow=TRUE), respect=TRUE)
plot(spr_birds~lat, d, pch=".", col=d$grp)
points(d$lat, predict(m2), cex=.3, col=d$wsm01+1)
legend("topright", cex=.7, levels(d$grp), col=1:length(levels(d$grp)), pch=16)
legend("topleft", c("wsm==0", "wsm>0"), col=1:2, lty=1)
```








# Residuals

We try to explain model residuals by `wsm`

```{r residuals}
for (area2 in levels(d$area2)) {
  sub_d = d[d$area2==area2 & d$wsm>0,]
  m = lm(spr_birds~lat, sub_d)
  layout(matrix(1:2, 1, byrow=TRUE), respect=TRUE)  
  plot(sub_d$lat, sub_d$spr_birds, pch=".", main=area2)
  points(sub_d$lat, predict(m), cex=.3)
  smoothScatter(m$residuals, sub_d$wsm)
  # mres = lm(sub_d$wsm~m$residuals)
  # abline(mres)
}
```


# Session Information

```{r results="verbatim"}
sessionInfo()
```




