---
title: "Modèles mixtes"
subtitle: "TP2: `biodivcontinamerdb`"
author: "Florent Chuffart & Laurent Husson"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---






```{r, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=4.5, eval=TRUE, echo=FALSE, results="hide")
```

# Data

  -  *lon*            : longitude
  -  *lat*            : lattitude
  -  *z*              : altitude
  -  *rug*            : rugosity
  -  *wsm*            : World Strain Map [1, Fig. 2a], a worldwide quantification of the rate at which the surface of the Earth*s crust is currently shortened, sheared or stretched by tectonic forces
  -  *spr_amphibians* : species richness for amphibians
  -  *spr_birds*      : species richness for birds   
  -  *spr_mammals*    : species richness for mammals
  -  *area*           : geographique area
                 
                 
                


```{r db_files, echo=TRUE, results="verbatim"}
biodivcontinamerdb = readRDS("biodivcontinamerdb.rds")
# set.seed(1)
# biodivcontinamerdb = biodivcontinamerdb[sort(sample(1:nrow(biodivcontinamerdb), 30000)), ]
# saveRDS(biodivcontinamerdb, "biodivcontinamerdb.rds")
# table(biodivcontinamerdb$area)
head(biodivcontinamerdb)
dim(biodivcontinamerdb)
```


```{r map, fig.height=6, fig.width=8}
d = biodivcontinamerdb
# layout(cbind(matrix(1, nrow=3, ncol=3), matrix(2:4)), respect=TRUE)
layout(1, respect=TRUE)
plot(  d$lon, d$lat, pch=".", col=c("grey75", "grey25")[(d$wsm>0)+1], main="North, Central and South American west coast" )
points(d$lon, d$lat, pch=".", col=adjustcolor(as.numeric(d$area), alpha.f=.1))
legend("topright", c("wsm==0", "wsm>0"), col=c("grey75", "grey25"), pch=15)
legend("bottomleft", levels(d$area), col=1:length(levels(d$area)), pch=15)

# y = unique(d$lat)
# x = sapply(y, function(lat){
#   tmpd = d[d$lat==lat,]
#   tmpd[which(tmpd$z==max(tmpd$z))[1],]$lon
# })
#
# x = x[order(y)]
# y = y[order(y)]
# lines(x,y, col=2)
#
layout(1, respect=TRUE)
plot(  d$lon, d$lat, pch=".", col=c("grey75", "grey25")[(d$wsm>0)+1], main="North, Central and South American west coast" )
tmpd = d[d$area == "sam_central" & d$spr_birds > 200,]
points(tmpd$lon, tmpd$lat, col=2)
tmpd = d[d$area == "sam_central" & d$spr_birds < 200,]
points(tmpd$lon, tmpd$lat, col=4)

# Warning: take care of factors!
# d$area2 = d$area
# d[d$area == "sam_central" & d$spr_birds > 200,]$area2 = "amazonia"
# table(d$area2 , useNA="ifany")

d$area2 = as.character(d$area)
d[d$area == "sam_central" & d$spr_birds > 200,]$area2 = "amazonia"
d$area2 = factor(d$area2, levels=c(levels(d$area), "amazonia"))
```


```{r spr, fig.height=3, fig.width=9}
layout(matrix(1:3, nrow=1), respect=TRUE)
plot(d$lat, d$spr_amphibians, xlab="lat", ylab="spr", main="amphibians", pch=".", col=adjustcolor(as.numeric(d$area), alpha.f=.1))
plot(d$lat, d$spr_birds     , xlab="lat", ylab="spr", main="birds"     , pch=".", col=adjustcolor(as.numeric(d$area), alpha.f=.1))
plot(d$lat, d$spr_mammals   , xlab="lat", ylab="spr", main="mammals"   , pch=".", col=adjustcolor(as.numeric(d$area), alpha.f=.1))
```





# Questions

1. Does the World Strain Map impact the species richness given the lattitude for birds?

2. We are used to saying that 1000m of elevation corresponds to a loss of biodiversity equivalent to 1000km towards the nearest pole [2]. According to the dataset, is this assertion right?







# Models



```{r, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=TRUE, results="verbatim", warning=FALSE)
```



```{r data, eval=TRUE, echo=TRUE, results="verbatim"}
# d = biodivcontinamerdb
d$grp = d$area
# d$grp = d$area2
# class
class(d) = c("nfnGroupedData", "nfGroupedData", "groupedData", "data.frame")
attributes(d)$formula = formula(spr_birds ~ lat | grp)
                                       


# layout(1, respect=TRUE)
# plot(spr_birds~lat, d, pch=".", col=d$grp)
# head(d[,-1])
# table(d$grp)
```




```{r lm}
m0 = lm(spr_birds~lat, d)
m1 = lm(spr_birds~lat+grp, d)
m2 = lm(spr_birds~lat*grp, d)
```


## `m0`
```{r echo=FALSE, results="verbatim"}
m0 = lm(spr_birds~lat, d)
summary(m0)
layout(matrix(1:2, 1, byrow=TRUE), respect=TRUE)
plot(spr_birds~lat, d, pch=".", col=d$grp)
abline(m0)
```

```{r echo=TRUE, results="verbatim"}
d$abslat = abs(d$lat)
m01 = lm(spr_birds~abslat+wsm, d)
layout(1, respect=TRUE)
plot(spr_birds~lat, d, pch=".", col=d$grp)
points(d$lat, predict(m01), cex=.3)
summary(m01)
```

```{r echo=TRUE, results="verbatim"}
d$wsm01 = (d$wsm != 0)+0
m03 = lm(spr_birds~abslat+wsm01, d)
layout(1, respect=TRUE)
plot(spr_birds~lat, d, pch=".", col=d$grp)
points(d$lat, predict(m03), cex=.3)
summary(m03)
```


```{r echo=TRUE, results="verbatim"}
d$abslat = abs(d$lat)
m02 = lm(spr_birds~abslat+z, d)
layout(1, respect=TRUE)
plot(spr_birds~lat, d, pch=".", col=d$grp)
points(d$lat, predict(m02), cex=.3)
summary(m02)
```

1 deg of lat. corresponds to 100km


```{r echo=TRUE, results="verbatim"}
print(m02$coef[2] * 10) 
print(m02$coef[3] * 1000) 
```

## `m1`
```{r echo=FALSE, results="verbatim"}
m1 = lm(spr_birds~lat+grp, d)
# m1 = lm(spr_birds~lat+sex+race_ethnicity, d)
summary(m1)
layout(matrix(1:2, 1, byrow=TRUE), respect=TRUE)
plot(spr_birds~lat, d, pch=".", col=d$grp)
abline(a=m1$coef[1]             , b=m1$coef[2], col=1)
n = length(levels(d$grp))
for (i in 2:n) {
  abline(a=m1$coef[1] + m1$coef[i+1], b=m1$coef[2], col=i)  
}
legend("topright", cex=.7, levels(d$grp), col=1:length(levels(d$grp)), pch=16)
```

## `m2`
```{r echo=FALSE, results="verbatim"}
m2 = lm(spr_birds~lat*grp, d)
summary(m2)
layout(matrix(1:2, 1, byrow=TRUE), respect=TRUE)
plot(spr_birds~lat, d, pch=".", col=d$grp)
abline(a=m2$coef[1]             , b=m2$coef[2]              , col=1)
n = length(levels(d$grp))
for (i in 2:n) {
  abline(a=m2$coef[1] + m2$coef[i+1], b=m2$coef[2] + m2$coef[n+i] , col=i)
}
legend("topright", cex=.7, levels(d$grp), col=1:length(levels(d$grp)), pch=16)
```


## `m3`
```{r echo=TRUE, results="verbatim"}
m3 = lme4::lmer(spr_birds ~ lat + (lat | grp), data=d)
summary(m3)
coef(m3)
layout(matrix(1:2, 1, byrow=TRUE), respect=TRUE)
plot(spr_birds~lat, d, pch=".", col=d$grp)
n = length(levels(d$grp))
for (i in 1:n) {
  abline(a=coef(m3)$grp[i,1], b=coef(m3)$grp[i,2], col=i)
}
legend("topright", cex=.7, levels(d$grp), col=1:length(levels(d$grp)), pch=16)

m31 = lme4::lmer(spr_birds ~ lat + (lat | grp) + (lat | wsm01), data=d)
plot(spr_birds~lat, d, pch=".", col=4)
points(d$lat, predict(m31), cex=.3, col=d$wsm01+1)
legend("topright", c("wsm=0", "wsm!=0"), col=1:2, pch=16)
summary(m31)
coef(m31)


m32 = lme4::lmer(spr_birds ~ lat + wsm01 + (lat + wsm01 | grp), data=d)
plot(spr_birds~lat, d, pch=".", col=4, main="m31")
points(d$lat, predict(m31), cex=.3, col=d$wsm01+1)
legend("topright", c("wsm=0", "wsm!=0"), col=1:2, pch=16)
plot(spr_birds~lat, d, pch=".", col=4, main="m32")
points(d$lat, predict(m32), cex=.3, col=d$wsm01+1)
legend("topright", c("wsm=0", "wsm!=0"), col=1:2, pch=16)
summary(m32)
coef(m32)


layout(matrix(1:2, 1, byrow=TRUE), respect=TRUE)

m32 = lme4::lmer(spr_birds ~ lat + wsm01 + (lat * wsm01 | grp), data=d)
plot(spr_birds~lat, d, pch=".", col=4, main="m32")
points(d$lat, predict(m32), cex=.3, col=d$wsm01+1)
legend("topright", c("wsm=0", "wsm!=0"), col=1:2, pch=16)
summary(m32)
coef(m32)

m33 = lme4::lmer(spr_birds ~ lat + (1 | grp) + (0 + lat | grp/wsm01), data=d)
plot(spr_birds~lat, d, pch=".", col=4, main="m33")
points(d$lat, predict(m33), cex=.3, col=d$wsm01+1)
legend("topright", c("wsm=0", "wsm!=0"), col=1:2, pch=16)
summary(m33)
coef(m33)

```

```{r}
m39 = lme4::lmer(spr_birds ~ abslat + z + (abslat + z | grp), data=d)
plot(spr_birds~lat, d, pch=".", col=4, main="m31")
points(d$lat, predict(m39), cex=.3, col=d$wsm01+1)
coef(m39)
cbind(
  coef(m39)$grp$abslat *10,
  coef(m39)$grp$z *1000
)

```

```


## `m4`
```{r echo=FALSE, results="verbatim"}
layout(matrix(1:2, 1, byrow=TRUE), respect=TRUE)
m4 = lme4::lmer(spr_birds ~ lat + (1 | grp), data = d)
summary(m4)
coef(m4)
plot(spr_birds~lat, d, pch=".", col=d$grp)
n = length(levels(d$grp))
for (i in 1:n) {
  abline(a=coef(m4)$grp[i,1], b=coef(m4)$grp[i,2], col=i)
}
legend("topright", cex=.7, levels(d$grp), col=1:length(levels(d$grp)), pch=16)
```


## `m5`
```{r echo=FALSE, results="verbatim"}
layout(matrix(1:2, 1, byrow=TRUE), respect=TRUE)
m5 = lme4::lmer(spr_birds ~ lat + (0+lat | grp), data = d)
summary(m5)
coef(m5)
plot(spr_birds~lat, d, pch=".", col=d$grp)
n = length(levels(d$grp))
for (i in 1:n) {
  abline(a=coef(m5)$grp[i,1], b=coef(m5)$grp[i,2], col=i)
}
legend("topright", cex=.7, levels(d$grp), col=1:length(levels(d$grp)), pch=16)
```




```{r}
BIC(m0)
BIC(m1)
BIC(m2)
# BIC(m3)
BIC(m3)
# BIC(m5)
BIC(m4)
# BIC(m7)
BIC(m5)

layout(matrix(1, 1, byrow=TRUE), respect=TRUE)

plot(
  0:5,
  c(
  BICm0=BIC(m0),
  BICm1=BIC(m1),
  BICm2=BIC(m2),
  BICm3=BIC(m3),
  BICm4=BIC(m4),
  BICm5=BIC(m5)
), ylab="BIC", xlab="models")

```







# References

[1] Kreemer, C., Blewitt, G., & Klein, E.C. (2014). *A geodetic plate motion and Global Strain Rate Model*. **Geochemistry, Geophysics, Geosystems**, 15(10), 3849-3889.

[2] *Un monde qui fond*, Camille Belsoeur, Arkhé 2022.

# Session Information

```{r results="verbatim"}
sessionInfo()
```




