---
title: "biodiv"
author: "Florent Chuffart & Laurent Husson"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---



```{r, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=4.5, eval=TRUE, echo=FALSE, results="hide")
```

# Question

Does the World Strain Map impact the species richness given the lattitude for amphibians? for birds? for mammals?

# Data

  -  *lon*            : longitude
  -  *lat*            : lattitude
  -  *z*              : altitude
  -  *rug*            : rugosity
  -  *wsm*            : World Strain Map (Kreemer et al., 2014, Fig. 2a), a worldwide quantification of the rate at which the surface of the Earth*s crust is currently shortened, sheared or stretched by tectonic forces
  -  *spr_amphibians* : species richness for amphibians
  -  *spr_birds*      : species richness for birds   
  -  *spr_mammals*    : species richness for mammals
  -  *area*           : geographique area
                 
                 
                


```{r db_files, echo=TRUE, results="verbatim"}
biodivcontinamerdb = readRDS("biodivcontinamerdb.rds")
head(biodivcontinamerdb)
dim(biodivcontinamerdb)
```


```{r fig.height=6, fig.width=8}
d = biodivcontinamerdb
# layout(cbind(matrix(1, nrow=3, ncol=3), matrix(2:4)), respect=TRUE)
layout(1, respect=TRUE)
plot(  d$lon, d$lat, pch=".", col=c("grey75", "grey25")[(d$wsm>0)+1], main="North, Central and South American west coast" )
points(d$lon, d$lat, pch=".", col=adjustcolor(as.numeric(d$area), alpha.f=.05))
legend("topright", c("wsm==0", "wsm>0"), col=c("grey75", "grey25"), pch=15)
legend("bottomleft", levels(d$area), col=1:length(levels(d$area)), pch=15)
```


```{r fig.height=3, fig.width=9}
layout(matrix(1:3, nrow=1), respect=TRUE)
plot(d$lat, d$spr_amphibians, xlab="lat", ylab="spr", main="amphibians", pch=".", col=adjustcolor(as.numeric(d$area), alpha.f=.1))
plot(d$lat, d$spr_birds     , xlab="lat", ylab="spr", main="birds"     , pch=".", col=adjustcolor(as.numeric(d$area), alpha.f=.1))
plot(d$lat, d$spr_mammals   , xlab="lat", ylab="spr", main="mammals"   , pch=".", col=adjustcolor(as.numeric(d$area), alpha.f=.1))
```


# References

Kreemer, C., Blewitt, G., & Klein, E.C. (2014). A geodetic plate motion and Global Strain Rate Model. Geochemistry, Geophysics, Geosystems, 15(10), 3849-3889.

# Session Information

```{r results="verbatim"}
sessionInfo()
```




